Color- and piece-blind chess

### 2019-03-14 chess/papers/blind.tex
Actually add Spycheck player, some writing, go to bed

Mainly, we need to store the pieces and their locations. There are a
few other bits, like whose turn it is (1 bit), whether each player can
still castle king- and queen-side (4 bits), and whether and where an
en passant capture is possible (4 bits).\footnote{
%
Technically, we need to store a lot of additional information with the
board in order to completely implement the rules of
chess.\cite{fiderules} The trickiest of these involve the rules for
draw by repetition, which make reference to the history of the game
(See Footnote~1 in {\it Survival in chessland}\cite{survival} and seem
to require storing all previous board states. Fortunately, if we are
being this picky, then we also know that the length of a chess game is
bounded by a constant: Rule 9.6.2 ends the game in a draw if both
players have made 75 moves without a pawn move or
capture,\footnotemark\ so it suffices to store the $75\times 2$ most recent
(half-)moves. This sucks so most people don't do it (for example, FEN
notation only gives the number of such moves, and so cannot implement
the draw by repetition). On the other hand, if we insist, then this
may give us a simpler route to a constant-space exhibition, since the
$B \stepsto{m} B'$ relation is probably reversible with such
information.}\footnotetext{These are two types of moves that make it impossible
  to formally repeat a position that preceded them. Castling also has
  this property, but doesn't count because it is a ``secret move.''}

With 64 squares, six pieces for each of the two sides, plus the empty
square, a straightforward representation of the board uses $64 \times 4=256$
bits. The Thursd'z Institute considered more compact
representations\cite{representations}; one nice choice works as follows:

Start with a single 64-bit mask which indicates, for each square on
the board, whether it contains any piece. Note that there can only be
up to 32 pieces on the board. To tell what these pieces are, we then
follow with 32 four-bit records; these indicate the piece's color and
type.\footnote{Since only 32 of the 64 bits can be set, you could do
  slightly better by representing $\binom{60}{32}$ in $\sim 61$ bits.
  When fewer than 32 squares are occupied, we can use a record
  containing e.g. a third king (otherwise impossible) to indicate that
  we should ignore the remaining bits. However, this gets vastly more
  complicated for only 3 bits of savings.} 
%
With the 9 bits of extra state above, this totals $64+32 \times 4+9=201$
bits. There is some slack in the representation, because there are
only 12 actual possibilities for a piece but we can indicate 16 with 4
bits. It is great to abuse this slack to save bits; for example, we
can store a new type of rook that is still able to castle (it can only
be in the corners and thus also indicates its own color), eliminating
the 4 castling bits. We can similarly introduce an en-passantable
pawn, saving the 4 bits of en passant state; this piece can only be in
ranks 4 or 5, so it also indicates its color. We can also have a
``king whose turn it is'' for each side, saving the side-to-move bit.
This totals a nice round $64+32 \times 4=192$ bits.\footnote{ Since this is
  SIGBOVIK, I am freed from the burden of comparing related work. I
  did however read the rather bad Wikipedia article on the
  topic\cite{wikipediaboard} which describes a Huffman-based encoding
  that uses a ``maximum of 204 bits, and often much less.'' This also
  includes a 7-bit 50-move counter (but you really need to implement a
  75-move counter; 50 moves only allow a player to {\em claim} a draw)
  so should be compared as $197$ bits. But the article also contains
  many bugs, like the misconception that there can only be four total
  rooks (pawns are allowed to promote to rook). So the approach
  described here is both {\em more efficient} and {\em more correct}.}
%
This would allow approximately an 11 billion-game simultaneous
exhibition in RAM on my desktop computer.

@@

Everything was trained on a single GPU, though a fairly decent one in
2018, the EVGA GeForce GTX 1080 FTW (really), using OpenCL. Biases
were initialized to $0$ and weights with a Gaussian of mean $0$ and
standard deviation $0.025$. In the first phase, there were 64 examples
per round, and after vacuuming, 2048. The round learning rate
$\alpha_r$ started at $0.1$ and descended linearly to $0.002$ over
500,000 rounds; and the learning rate when updating weights (for each
example) $\alpha$ is $\alpha_r / \verb+examples_per_round+$. In no way
am I recommending these parameters. Fiddling with the parameters to
make it do its black magic or alternately carom off to a sea of NaNs
or zeroes is for sure the worst part of neural networks. In fact, I
initially started with the classical sigmoid transfer function, but
``upgraded'' to the ``leaky rectified linear'' function
%
$$(p < 0) \,?\, p \times 0.01 : p$$
%
after getting fed up with sigmoid weights caroming off (see
``vanishing gradient problem'' and/or ``exploding gradient problem'').
The final model was trained over 339,885 rounds on 223 million
examples. It did not appear to show signs of improving for
several days before I terminated it.

@@

\begin{figure}[ht]
XXX actual example
\caption{The white queen ``disappears'' after QxWhatever, because
  unblinding predicts it to be one of black's own pawns, which is far
  more likely in that square.} \label{fig:trick}
\end{figure}

A few things are distinctly disappointing about its performance. Even
outside of ``likely'' positions, it usually predicts that pieces on
black's side of the board are black, and vice versa
(Figure~\ref{fig:allon}). This makes sense, but suggests serious
limitation on using the prediction to play chess. Less forgivably, it
sometimes predicts more than one king per side (or zero), which is
always wrong. Actually, an early version had this problem in spades,
usually frequently predicting two or three kings. Upon debugging, I
had simply made the noob mistake of printing both King and Knight with
the letter ``K.'' Ha! It often predicts the ``wrong'' number of
bishops (etc.), or places them on the same color. This is technically
possible through promotion, but usually a bad guess, since promotion
is relatively rare, and moreso promotion to a piece other than a
queen. An approach that might handle this better (but may have
different downsides) would be instead to predict the ``fates'' of the
32 initial pieces~\cite{survival}. The fate of a pawn includes what if
any piece it has promoted to, but no other pieces can. This would {\em
  require} that the model only predict a single location for each
king, among other things. However, this would require a much larger
output layer (32 pieces can move to 64 squares, plus promotion) and it
is not always clear how to interpret its value into a position (for
example, if two pieces are predicted strongly to be on the same
square).

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{blind-allon}
\caption{What the model predicts (in single-king mode;
  Section~\ref{sec:using}) for {\tt 0xFFFFFFFFFFFFFFFF}, the board
  with all bits set. This is an impossible position, but it gives some
  idea of the model's biases for each square. Notably, most of the pieces
  on each half of the board have a single color. This makes sense, but
  also suggests substantial limitation. When single-king mode is off,
  the bottom right king is predicted as a white rook.} \label{fig:allon}
\end{figure}

@@

Once we have predicated a board state, we can play with it. The
simplest way to do this is to use a strong chess engine to pick a move
for the position. Here, I use Stockfish with a node budget of 1
million nodes, which takes about 1 second of CPU time per move on my
computer; henceforth {\sf stockfish1m}.

There are some complications:
\begin{itemize}
\item Frequently, the unblinded board will not match the true
  position, and Stockfish will choose a move that is illegal. So, as discussed
  before, we actually return a prioritzed list of moves. For this first
  experiment, we just return the move Stockfish recommends, followed by all
  other moves ordered randomly.
\item Stockfish is a very strong engine, and in my opinion the code is generally
  good, but it is very sensitive to bad FEN (the notation used to give it a
  position to evaluate) strings. Given a bad string, like one that says castling
  is possible when the king isn't in its home square, often crashes the engine.
  So we need to make sure to actually pass valid positions. I accomplish this
  by making the following modifications to the predicted board:
  \begin{itemize}
  \item If a castling bit is set, but castling is not possible due to the position
    of the king or rook, I clear the bit.
  \item Set the side-to-move to be the correct actual value. This uses
    the unblinded state, so is superficially cheating. But note that
    if we get the side wrong, then Stockfish's move will always be
    illegal: Moves are specified as a source and destination
    square,\footnote{Plus promotion piece. Castling is represented as
      a two-square move of the king.} and so the source square of
    Stockfish's move would always be a piece of the wrong piece's
    color. So this is equivalent to (but twice as efficient as)
    running stockfish twice, one for each side, and prioritizing the
    predicted side's move first.
  \end{itemize}
  This won't fix all positions, for example, if the white and black king are
  adjacent to one another in mutual check. If an irreparable problem is detected,
  then I just return a uniformly random move list.
\end{itemize}

It is easy to beat this chess engine, by tricking it as in
Figure~\ref{fig:trick}, although this involves unnatural moves, so it
may only apply if you know how it works.
% XXX example game?
Measuring how well it plays in an absolute sense is a subject of some
interest, so I wrote a separate paper about that~\cite{eloworld}. This
algorithm, called \verb+blind_yolo+, had an Elo World score of XXX.

\subsection{We three kings}

@@

The predicted board often expresses uncertainty about some squares,
which could be thought of as probabilities. A principled improvement
would be to try to find moves that are good in expectation, that is,
integrated all possible boards in proportion to their predicted
probability. A good approximation might be had by sampling a bunch
of boards according to the predicted distribution, and then using
Stockfish to score the top $k$ moves for each; we can then order
moves by their expected score. Unfortunately, it is not easy to
efficiently get Stockfish to generate scored moves for $k \neq 1$.
Even with $k = 1$, this approach would be slow, taking about a
second for each (distinct) sampled board. So I did not try it,
at least not before submitting this paper.

\medskip
I initially considered trying to solve this whole problem with neural
networks. The current best known engine in the world (AlphaZero) at
least {\em uses} a neural network.
% XXX is the whole thing just a network? I guessed it's probably
% alpha-beta plus a NN evaluation function?
The biggest advantage would be that it would naturally be able to
consider multiple moves under uncertainty about the board state, as
just discussed, without any particular extra logic. My plan was to
make a multiple different components that coud be evaluated
separately, starting with the unblinder described, followed by a unit
that predicts legal moves, and then a unit that takes these two (and
also the 64-bit blinded mask if it likes) and scores each move.
Training a reasonable unblinder took longer than I expected, and the
legal move predictor never really worked that well,
% (XXX some evaluation of it)
so I just settled for basing it off the single unblinder unit. Can you
do better?

@@

I would like to thank the little people (pawns) and the anonymous
referrees.

>>>>

Mainly, we need to store the pieces and their locations. There are a
few other bits, like whose turn it is (1 bit), whether each player can
still castle king- and queen-side (4 bits), and whether and where an
en passant capture is possible (4 bits).\footnote{
%
Technically, we need to store a lot of additional information with the
board in order to completely implement the rules of
chess.\cite{fiderules} The trickiest of these involve the rules for
draw by repetition, which make reference to the history of the game
(See Footnote~1 in {\it Survival in chessland}~\cite{survival}) and seem
to require storing all previous board states. Fortunately, if we are
being this picky, then we also know that the length of a chess game is
bounded by a constant: Rule 9.6.2 ends the game in a draw if both
players have made 75 moves without a pawn move or
capture,\footnotemark\ so it suffices to store the $75\times 2$ most recent
(half-)moves. This sucks so most people don't do it (for example, FEN
notation only gives the number of such moves, and so cannot implement
the draw by repetition). On the other hand, if we insist, then this
may give us a simpler route to a constant-space exhibition, since the
$B \stepsto{m} B'$ relation is probably reversible with such
information.}\footnotetext{These are two types of moves that make it impossible
  to formally repeat a position that preceded them. Castling also has
  this property, but doesn't count because it is a ``secret move.''}

With 64 squares, six pieces for each of the two sides, plus the empty
square, a straightforward representation of the board uses $64 \times 4=256$
bits. The Thursd'z Institute considered more compact
representations~\cite{representations}; one nice choice works as follows:

Start with a single 64-bit mask which indicates, for each square on
the board, whether it contains any piece. Note that there can only be
up to 32 pieces on the board. To tell what these pieces are, we then
follow with 32 four-bit records; these indicate the piece's color and
type.\footnote{Since only 32 of the 64 bits can be set, you could do
  slightly better by representing $\binom{64}{32}$ in $\sim 61$ bits.
  When fewer than 32 squares are occupied, we can use a record
  containing e.g. a third king (otherwise impossible) to indicate that
  we should ignore the remaining bits. However, this gets vastly more
  complicated for only 3 bits of savings.} 
%
With the 9 bits of extra state above, this totals $64+32 \times 4+9=201$
bits. There is some slack in the representation, because there are
only 12 actual possibilities for a piece but we can indicate 16 with 4
bits. It is great to abuse this slack to save bits; for example, we
can store a new type of rook that is still able to castle (it can only
be in the corners and thus also indicates its own color), eliminating
the 4 castling bits. We can similarly introduce an en-passantable
pawn, saving the 4 bits of en passant state; this piece can only be in
ranks 4 or 5, so it also indicates its color. We can also have a
``king whose turn it is'' for each side, saving the side-to-move bit.
This totals a nice round $64+32 \times 4=192$ bits.\footnote{ Since this is
  SIGBOVIK, I am freed from the burden of comparing related work. I
  did however read the rather bad Wikipedia article on the
  topic~\cite{wikipediaboard} which describes a Huffman-based encoding
  that uses a ``maximum of 204 bits, and often much less.'' This also
  includes a 7-bit 50-move counter (but you really need to implement a
  75-move counter; 50 moves only allow a player to {\em claim} a draw)
  so should be compared as $197$ bits. But the article also contains
  many bugs, like the misconception that there can only be four total
  rooks (pawns are allowed to promote to rook). So the approach
  described here is both {\em more efficient} and {\em more correct}.}
%
This would allow approximately an 11 billion-game simultaneous
exhibition in RAM on my desktop computer.

@@

Everything was trained on a single GPU, though a fairly decent one in
2018, the EVGA GeForce GTX 1080 ``FTW'' (really), using OpenCL. Biases
were initialized to $0$ and weights with a Gaussian of mean $0$ and
standard deviation $0.025$. In the first phase, there were 64 examples
per round, and after vacuuming, 2048. The round learning rate
$\alpha_r$ started at $0.1$ and descended linearly to $0.002$ over
500,000 rounds; and the learning rate when updating weights (for each
example) $\alpha$ is $\alpha_r / \verb+examples_per_round+$. In no way
am I recommending these parameters. Fiddling with the parameters to
make it do its black magic or alternately carom off to a sea of NaNs
or zeroes is for sure the worst part of neural networks. Indeed, I
initially started with the classical sigmoid transfer function, but
``upgraded'' to the ``leaky rectified linear'' function
%
$$(p < 0) \,?\, p \times 0.01 : p$$
%
after getting fed up with sigmoid weights caroming off (see
``vanishing gradient problem'' and/or ``exploding gradient problem'').
The final model was trained over 339,885 rounds on 223 million
examples. It did not appear to show signs of improving for
several days before I terminated it.

@@

\begin{figure}[ht]
\chessboard[tinyboard,setfen=rnbqkbnr/ppp2pQp/8/3pp3/4P3/8/PPPP1PPP/RNB1KBNR b KQkq - 0 3,showmover=false] (a)
  \includegraphics[width=0.75 \linewidth]{queendisappears} (b)
  \caption{(a) Position after 1.~e4 e5 2.~Qg4 d5 3.~Qxg7; note the white
    queen strangely on g7.
    (b) The bitmask for this position and the unblinder's prediction.
    The queen ``disappears'' after Qxg7, because
    unblinding predicts it to be one of black's own pawns---far
    more likely in that square.} \label{fig:trick}
\end{figure}

A few things are distinctly disappointing about its performance. Even
outside of ``likely'' positions, it usually predicts that pieces on
black's side of the board are black, and vice versa
(Figure~\ref{fig:allon}). This makes sense, but suggests serious
limitation on using the prediction to play chess. Less forgivably, it
sometimes predicts more than one king per side (or zero), which is
always wrong. Actually, an early version had this problem in spades,
frequently predicting two or three kings. Upon debugging, I had simply
made the noob mistake of printing both King and Knight with the letter
``K.'' Ha! It often predicts the ``wrong'' number of bishops (etc.),
or places them on the same color. This is technically possible through
promotion, but usually a bad guess, since promotion is relatively
rare, and moreso promotion to a piece other than a queen. An approach
that might handle this better (but may have different downsides) would
be instead to predict the ``fates'' of the 32 initial
pieces~\cite{survival}. The fate of a pawn includes what if any piece
it has promoted to, but this is not necessary for the other pieces.
This would {\em require} that the model only predict a single location
for each king, among other things. However, this would require a much
larger output layer (32 pieces can move to 64 squares, plus promotion)
and it is not always clear how to interpret its value into a position
(for example, if two pieces are predicted strongly to be on the same
square).

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{blind-allon}
\caption{What the model predicts (in single-king mode;
  Section~\ref{sec:singleking}) for {\tt 0xFFFFFFFFFFFFFFFF}, the board
  with all bits set. This is an impossible position, but it gives some
  idea of the model's biases for each square. Notably, most of the pieces
  on each half of the board have a single color. This makes sense, but
  also suggests substantial limitation. When single-king mode is off,
  the bottom right king is predicted as a white rook.} \label{fig:allon}
\end{figure}

@@

Once we have predicted a board state, we can play with it. The
simplest way to do this is to use a strong chess engine to pick a move
for the position. Here, I use Stockfish with a node budget of 1
million nodes, which takes about 1 second of CPU time per move on my
computer. There are some complications:
\begin{itemize}
\item Frequently, the unblinded board will not match the true
  position, and Stockfish will choose a move that is illegal. So, as discussed
  before, we actually return a prioritized list of moves. For this first
  experiment, we just return the move Stockfish recommends, followed by all
  other moves ordered randomly.
\item Stockfish is a very strong engine, and in my opinion the code is generally
  good, but it is very sensitive to bad FEN (the notation used to give it a
  position to evaluate) strings. Given a bad string, like one that says castling
  is possible when the king isn't in its home square, often crashes the engine.
  So we need to make sure to actually pass valid positions. I accomplish this
  by making the following modifications to the predicted board:
  \begin{itemize}
  \item If a castling bit is set, but castling is not possible due to
    the position of the king or rook, clear the bit.
  \item Set the side-to-move to be the correct actual value. This uses
    the unblinded state, so is superficially cheating. But note that
    if we get the side wrong, then Stockfish's move will always be
    illegal: Moves are specified as a source and destination
    square,\footnote{\,Plus promotion piece. Castling is represented as
      a two-square move of the king.} and so the source square of
    Stockfish's move would always be a piece of the wrong piece's
    color. So this is equivalent to (but twice as efficient as)
    running stockfish twice, one for each side, and prioritizing the
    predicted side's move first.
  \end{itemize}
  This won't fix all positions, for example, if the white and black king are
  adjacent to one another in mutual check. If an irreparable problem is detected,
  then I just return a uniformly random move list.
\end{itemize}

It is easy to beat this chess engine, by tricking it as in
Figure~\ref{fig:trick}, although this involves unnatural moves, so it
may only apply if you know how it works.
% XXX example game?
Measuring how well it plays in an absolute sense is a subject of some
interest, so I wrote a separate paper about that~\cite{eloworld}. This
algorithm, called \verb+blind_yolo+, had an Elo World score of XXX. It
beats a purely random player with a score of 101 wins, 27 losses, and
389 draws. Making moves purely at random is one of the few fair
comparisons, since the random strategy also works with color- and
piece-blind chess.

\subsection{We three kings} \label{sec:singleking}

@@

The predicted board often expresses uncertainty about some squares,
which could be thought of as probabilities. A principled improvement
would be to try to find moves that are good in expectation, that is,
integrated over all possible boards in proportion to their predicted
probability. A good approximation might be had by sampling a bunch of
boards according to the predicted distribution, and then using
Stockfish to score the top $k$ moves for each; we can then order moves
by their expected score. Unfortunately, it is not easy to efficiently
get Stockfish to generate scored moves for $k \neq 1$. Even with $k =
1$, this approach would be slow, taking about a second for each
(distinct) sampled board. So I did not try it, at least not before
submitting this paper.

\subsubsection{No, {\em u r} a lnetwork} \label{sec:legalmove}
I initially considered trying to solve this whole problem with neural
networks. The current best known engine in the world (AlphaZero) at
least {\em uses} a neural network.
% XXX is the whole thing just a network? I guessed it's probably
% alpha-beta plus a NN evaluation function?
The biggest advantage would be that it would naturally be able to
consider multiple moves under uncertainty about the board state, as
just discussed, without any particular extra logic. My plan was to
make multiple different components that could be evaluated
separately, starting with the unblinder described, followed by a unit
that predicts legal moves, and then a unit that takes these two (and
also the 64-bit blinded mask if it likes) and scores each move.
Predicting a legal move is also a natural function for machine
learning; a move can be given just as a source and destination
square.\footnote{ There are also four choices for promotion when
  moving a pawn into the last rank. It is always the case that if any
  promotion is legal, all choices are legal, so this does not need to
  be encoded in this phase. Also, at this level of play, always
  promoting to queen is a very safe simplification.} Many pairs of
squares are always impossible (e.g. no piece can ever move from A1 to
B8); so there are only 1792 potential moves to predict. However,
training a reasonable unblinder took longer than I expected, and the
legal move predictor never really worked that well (it has a harder
job),
% (XXX some evaluation of it)
so I just settled for basing it off the single unblinder unit. Can you
do better?

@@

I would like to thank the little people (pawns) and the the author-
and content-blind anonymous referrees.
